master thesis

### 2018-03-22 Paper - Latex/Paper - Skizze - v4.tex
Paper - Methods - update Plots; remove some subsections

\begin{document}
  % Titelseite
  \begin{titlepage}
    \pagestyle{empty}
  	\begin{center}
      {\Large Freie Universität Berlin} \\
    	\begin{Huge}
      	Fachbereich Mathematik und Informatik \\
      	\vspace{3mm}
    	\end{Huge}
    	\vspace{2cm}
    	\begin{Large}
        \textbf{An analysis of maximum parsimony algorithms to predict parasitism in Eukaryota} \\
        \vspace{3mm}
        using a large multifurcated phylogenetic synthesis tree \\
    	\end{Large}
      \vspace{3cm}
      \textbf{Eingereicht am:} \\
      3.4.2018 \\
    	\vspace{2cm}
    	Lydia Buntrock \\
      E-Mail: info@irallia.de \\
     	\vspace{3cm}
      \textbf{Betreuer:} \\
      Dr. Bernhard Y. Renard \\
      \& \\
      Prof. Dr. rer. nat. Emanuel Heitlinger \\      
  	\end{center}
    \clearpage
    \pagenumbering{Roman}
  \end{titlepage}

@@

%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%--------------------------------------------------------------------------------------------------- 
\chapter{Methods}
  As stated in the introduction, in this thesis, a maximum parsimony algorithm is applied to the 
    whole tree of life to obtain an ancestral state reconstruction of free-living versus parasite 
    states. \\
  So far, these reconstructions have been made mainly on binary trees with better data availability. 
    Therefore, a simulation is first performed to evaluate existing algorithms and decide how they 
    may be adapted to our given problem. This is to perform the ancestral state reconstruction for a 
    multifurcated (non-binary) tree using binary states. \\
  Accordingly, in addition to the necessary data sets (GloBI, OTL), the chosen algorithm and the 
    evaluation of its results, this chapter also deals with the previously performed simulation and 
    the evaluation of the various algorithms and their parameters. \\
  Figure \ref{fig:workflow} briefly outlines these relationships. A more detailed view of the 
    workflow can be found in the appendix \ref{fig:BigWorkflow}.
  \begin{figure}[h!]
    \centering
    \includegraphics[width=0.45\textwidth]{Figures/Workflow-overview.png}
    \caption{The Workflow of the resulting procedure with the following steps: \\
      (1) Retrieve phylogenentic tree data as input for the tree (OTL) and the state information (GloBI).
      (2) Get metadata of these for a realisitc the simulation of the maximum parsimony algorithms (Fitch \& Sankoff).
      (3) Build and run the simulation.
      (4) Evaluation of parameters for the simulation and the ancestral state reconstruction of the real tree.
      (5) Evaluate the accuracy of developed algorithms and choose the best.
      (6) Run the resulted algorithm on the original data.
      (7) Evalute and interprete results. %$\rightarrow$ Origins etc...
    }
    \label{fig:workflow}
  \end{figure}

@@

      For the maximum parsimony analysis of the real data, all transition probabilities are equated.
        However, the used castor package \cite{Louca2017} offers the possibility to enter different 
        transition probabilities.
      \begin{wrapfigure}{r}{0.5\textwidth}
        \begin{center}
          \includegraphics[trim = 0mm 0mm 0mm 13mm, clip, width=0.48\textwidth]{Figures/40-60.png}
        \end{center}
        \caption{60~\% Free-living - 40~\% Parasites \\ red: parasites, blue: free-living, \\ the threshold is at 0.4}
        \label{fig:Beta distribution}
      \end{wrapfigure}

      In the simulation two beta distributions have been chosen and a threshold that indicates the change 
        between states. \\
      Different thresholds with different beta distributions are simulated, with different distributions 
        of parasites and free-livings:
        \begin{itemize}
          \item 50~\% P to 50~\% FL,
          \item 40~\% P to 60~\% FL,
          \item 30~\% P to 70~\% FL and 
          \item 20~\% P to 80~\% FL
        \end{itemize}
        (\todo{see results simulation, ref...}). Figure \ref{fig:Beta distribution} 
        shows one example of these. \\
      \todo{Plot neu erstellen: Achsenbeschriftung, threshold}

@@

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{Fitch maximum parsimony}
      Fitch maximum parsimony is an algorithm for rooted, binary trees and describes an ancestral state 
        reconstruction for discrete states \cite{Fitch1971} by minimizing transitions between states. \\
      Note, the original Fitch algorithm has the sole purpose of minimizing the number of transitions 
        and not reconstructing the ancestral nodes. Felsenstein \cite{Felsenstein2003} describes a 
        simple extension for the reconstruction. Cunningham et al. \cite{Cunningham1998} have refined 
        these. \todotext{Wir haben mit ein paar kleinen änderungen optimiert... und schließlich auf 
        multifurcated angepasst...} \todotext{eigentlich ist Cunningham 'nur' eine kritische 
        Neubewertung. Sie beziehen ihren Algortihmus auf Swofford und Maddison...} \\
      To understand the differences to the multifurcated case, the algorithm for the binary case is 
        briefly explained and referred to the extension. \\
        \begin{wrapfigure}[19]{r}{0.5\textwidth}
          \centering
          \includegraphics[width=0.4\textwidth]{Figures/Fitch1.png}
          \caption{Fitch algorithm for binary trees. \\
            The unknown leaf node is discribed with both states. Computed internal nodes (exclusive the 
            root node) consists of three sets, where the last set is the final one (bold). \\
            From the second internal node (seen from the root node) there are several possibilities to 
            create the second and third set.}
          \label{fig: binary Fitch}
        \end{wrapfigure} 
      Input: A rooted, binary tree, with state informations in the leaf nodes. Each node is depicted as 
        a set of states. There are only two states in this thesis, free-living (FL) and parasitic (P). 
        Internal nodes have three sets, which are empty at the beginning, excluding the root node, it 
        has only one. Leaf nodes have their state as a set (eg \{FL\} or \{P\}, unknown leaf nodes the 
        union of all possible states (\{FL, P\}). \\

      The algorithm traverses three times trough the tree and fills these sets. \\
      In each step, two sets are considered and their intersect formed. There are two cases:
      \begin{enumerate}
        \item The intersection is not empty and corresponds to the new set.
        \item The intersection is empty. $\rightarrow$ Build the union of these sets as new set.
      \end{enumerate}
      First traverse from the leaf nodes to the root / move down the tree / postorder tree traversal.
        Each internal node is formed from its child nodes, where at the beginning the only information
        lies. \\
      Second traverse from the root node to the leafs. Each internal node is formed from its father node 
        and its sibling node. \\
      Last traversion (direction does not matter): Build the final state for every node. It is formed 
        from the sets of previous traversals. \\
      (The original Fitch algorithm was designed to minimize transitions without predicting actual states 
        of internal nodes, so it was just the first traversal.) \\
      The extension to the non-binary case is quite obvious, but holds some opportunities. In this case, 
        more than two children may be present for the first traversal, but the incision or union may 
        also be formed over more than two sets. Also in the second traversing, there may be several 
        sibling nodes. However, there are several possibilities here that were all tested and compared 
        in the simulation. Some of these options are already available in the binary case:
      \begin{itemize}
        \item The father node has (except for the root node) two state sets, because he came through 
          the up-traversing previously. Are both sets used or only the first traversing?
        \item Since there are several siblings, do \colorbox{red}{you} first of all make the cut or union, 
          or directly in the whole with the father node?
      \end{itemize}
      The first point already has an effect on the binary case. Figure \ref{fig: binary Fitch} shows 
        both possibilities of the three sets. \\
      Cunningham uses only the first state set of the father node \cite{Cunningham1998}. \\
      From these two points four different versions of Fitch were formed:
      \begin{enumerate}
        \item Fitch 1: First state set of father node; intersection/union of siblings first.
        \item Fitch 2: First state set of father node; intersection/union of siblings together with father node.
        \item Fitch 3: Both state sets of father node; intersection/union of siblings first.
        \item Fitch 4: Both state sets of father node; intersection/union of siblings together with father node sets.
      \end{enumerate}
      \begin{figure}
        \centering
        \includegraphics[width=0.8\textwidth]{Figures/simulation_fitch_evaluation.png}
        \caption{Test of Fitch Versions.}
        \label{fig:Fitch versions}
      \end{figure}
      These four versions were tested in the simulation with 100 trees and 10000 leaf nodes and a
        distribution of 60~\% FL to 40~\% P. Figure \ref{fig:Fitch versions} shows this over all unknown 
        node percentage. \\
      % At 90~\% unknown nodes and 90~\% of multifurcation of the internal nodes, version 1 was 89.26~\%, 
      %   version 2 was 89.26~\%, version 3 was 89.35~\%, and version 4 was 89.31~\% correct. Therefore, 
      %   only version 3 was used for all further simulations.
      At 95~\% unknown nodes and 95~\% of multifurcation of the internal nodes, version 1 was 88.37~\%, 
        version 2 was 88.37~\%, version 3 was 88.4~\%, and version 4 was 88.39~\% correct. Therefore, 
        only version 3 was used for all further simulations.

@@

    \begin{figure}[h!]
      \centering
      \includegraphics[width=1\textwidth]{Figures/Workflow-Simulation.png}
      \caption{Course of the simulation with influence of the metadata analysis from the real data.}
      \label{fig:Simulation Workflow}
    \end{figure}

    \todotext{Figure \ref{fig:Simulation Workflow} shows the course of the simulation. The individual 
      steps are explained in the following subsections: ...}

    \todo{Evaluation - compare trees (distances)}

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{random binary tree}
      A tree is needed to do a simulation of ancestral state reconstruction. It had to be decided whether 
        to take the real tree or simulate a tree. In this simulation, trees are created randomly, as 
        one can replicate a complete binary phylogentic tree. Thus, there is also the possibility to 
        simulate the multifurcation. \\

      To get a random binary tree, the Phylo package from biopython is used \cite{Cock2009}. They offer 
        a randomized function which returns a BaseTree\footnote{
          \hyperlink{https://github.com/biopython/biopython/blob/master/Bio/Phylo/BaseTree.py}
          {https://github.com/biopython/biopython/blob/master/Bio/Phylo/BaseTree.py}
        }. \\
      \todo{ref in die discussion über die randomized function? Diskutieren wir das?}

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{simulating states and transitions between them}
      The next step is to simulate the states of the nodes using the transitions. Again, fully known 
        states are simulated and then everything is 'forgotten' but a few in the leaf nodes so that 
        \colorbox{red}{you} can later compare the reconstruction with the origin. \\
      The root node is defined as ancestor of all subsequent species and in this case, determined to be
        free-living. Therefore, a beta distribution for free-living is used at the beginning. Now 
        traverse from the root to the leaf nodes, always pulling out of the current distribution until 
        \colorbox{red}{you} get above the threshold and the new node changes state. \anmerkungstext{Für 
        den Tree/Graph Traversal würde sich eine erklärende Abbildung anbieten. Nur Text macht es sehr 
        schwer verständlich an der Stelle. (Thilo)} \\
      To ensure that the parameter of the binomial distribution is restricted to the [0,1] interval, it 
        is modeled with a beta distribution as in Figure \ref{fig:Beta distribution}. \\

@@

% simulate loss of information
      Here begins the simulation of the lost information. This is on the one hand the states and on the 
        other the topology of the tree. Some splits of nodes are unknown with which the tree is 
        multifurcated (explained in the following section \todo{pageref}). \\

      In the real tree, there is usually only information about species living today $\rightarrow$ leaf 
        nodes. And beyond only a small percentage of these. All information about the states of the 
        internal node and one leaf node is 'forgotten' and stored in another column to the node. \\
      Different percentages of forgetting the information are simulated, as \colorbox{red}{you} can read 
      in the \todo{section ... from the results}.

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{simulating loss of information of the tree topology}
      As previously explained, some divisions in the tree are not known, so the real tree is not binary.
      This multifurcation was simulated by an equally distributed percentage of forgotten internal nodes.

>>>>

\begin{document}
  % Titelseite
  \begin{titlepage}
    \pagestyle{empty}
  	\begin{center}
      {\Large Freie Universität Berlin} \\
    	\begin{Huge}
      	Fachbereich Mathematik und Informatik \\
      	\vspace{3mm}
    	\end{Huge}
    	\vspace{2cm}
    	\begin{Large}
        \textbf{An analysis of maximum parsimony algorithms to predict parasitism in Eukaryota} \\
        \vspace{3mm}
        using a large multifurcated phylogenetic synthesis tree \\
    	\end{Large}
      \vspace{3cm}
      \textbf{Eingereicht am:} \\
      3.4.2018 \\
    	\vspace{2cm}
    	Lydia Buntrock \\
      E-Mail: info@irallia.de \\
     	\vspace{3cm}
      \textbf{Betreuer:} \\
      Prof. Dr. Bernhard Y. Renard \\
      \& \\
      Prof. Dr. rer. nat. Emanuel Heitlinger \\      
  	\end{center}
    \clearpage
    \pagenumbering{Roman}
  \end{titlepage}

@@

%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%--------------------------------------------------------------------------------------------------- 
\chapter{Methods}
  As stated in the introduction, in this thesis, a maximum parsimony algorithm is applied to the 
    whole tree of life to obtain an ancestral state reconstruction of free-living versus parasite 
    states. \\
  So far, these reconstructions have been made mainly on binary trees with better data availability. 
    Therefore, a simulation is first performed to evaluate existing algorithms and decide how they 
    may be adapted to our given problem. This is to perform the ancestral state reconstruction for a 
    multifurcated (non-binary) tree using binary states. \\
  Accordingly, in addition to the necessary data sets (GloBI, OTL), the chosen algorithm and the 
    evaluation of its results, this chapter also deals with the previously performed simulation and 
    the evaluation of the various algorithms and their parameters. \\
  Figure \ref{fig:workflow} briefly outlines these relationships. A more detailed view of the 
    workflow can be found in the appendix \ref{fig:BigWorkflow}.
  \begin{figure}[h!]
    \centering
    \includegraphics[width=0.6\textwidth]{Figures/Workflow-overview.png}
    \caption{The Workflow of the resulting procedure with the following steps: \\
      (1) Retrieve phylogenentic tree data as input for the tree (OTL) and the state information (GloBI).
      (2) Get metadata of these for a realisitc the simulation of the maximum parsimony algorithms (Fitch \& Sankoff).
      (3) Build and run the simulation.
      (4) Evaluation of parameters for the simulation and the ancestral state reconstruction of the real tree.
      (5) Evaluate the accuracy of developed algorithms and choose the best.
      (6) Run the resulted algorithm on the original data.
      (7) Evalute and interprete results. %$\rightarrow$ Origins etc...
    }
    \label{fig:workflow}
  \end{figure}

@@

      For the maximum parsimony analysis of the real data, all transition probabilities are equated.
        However, the used castor package \cite{Louca2017} offers the possibility to enter different 
        transition probabilities.
      \begin{wrapfigure}{r}{0.5\textwidth}
        \begin{center}
          \includegraphics[trim = 0mm 0mm 0mm 0mm, clip, width=0.45\textwidth]{Figures/40-60.JPG}
        \end{center}
        \caption{60~\% Free-living - 40~\% Parasites \\ red: parasites, blue: free-living, \\ the threshold is at 0.4}
        \label{fig:Beta distribution}
      \end{wrapfigure}

      In the simulation two beta distributions have been chosen and a threshold that indicates the change 
        between states. \\
      Different thresholds with different beta distributions are simulated, with different distributions 
        of parasites and free-livings:
        \begin{itemize}
          \item 50~\% P to 50~\% FL,
          \item 40~\% P to 60~\% FL,
          \item 30~\% P to 70~\% FL and 
          \item 20~\% P to 80~\% FL
        \end{itemize}
        (\todo{see results simulation, ref...}). Figure \ref{fig:Beta distribution} 
        shows one example of these. \\

@@

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{Fitch maximum parsimony}
      Fitch maximum parsimony is an algorithm for rooted, binary trees and describes an ancestral state 
        reconstruction for discrete states \cite{Fitch1971} by minimizing transitions between states. \\
      Note, the original Fitch algorithm has the sole purpose of minimizing the number of transitions 
        and not reconstructing the ancestral nodes. Felsenstein \cite{Felsenstein2003} describes a 
        simple extension for the reconstruction. Cunningham et al. \cite{Cunningham1998} have refined 
        these. \todotext{Wir haben mit ein paar kleinen änderungen optimiert... und schließlich auf 
        multifurcated angepasst...} \todotext{eigentlich ist Cunningham 'nur' eine kritische 
        Neubewertung. Sie beziehen ihren Algortihmus auf Swofford und Maddison...} \\
      To understand the differences to the multifurcated case, the algorithm for the binary case is 
        briefly explained and referred to the extension. \\

      Input: A rooted, binary tree, with state informations in the leaf nodes. Each node is depicted as 
        a set of states. There are only two states in this thesis, free-living (FL) and parasitic (P). 
        Internal nodes have three sets, which are empty at the beginning, excluding the root node, it 
        has only one. Leaf nodes have their state as a set (eg \{FL\} or \{P\}, unknown leaf nodes the 
        union of all possible states (\{FL, P\}). \\

      The algorithm traverses three times trough the tree and fills these sets. \\
      In each step, two sets are considered and their intersect formed. There are two cases:
      \begin{enumerate}
        \item The intersection is not empty and corresponds to the new set.
        \item The intersection is empty. $\rightarrow$ Build the union of these sets as new set.
      \end{enumerate}
      \begin{wrapfigure}{r}{0.5\textwidth}
        \centering
        \includegraphics[width=0.4\textwidth]{Figures/Fitch1.png}
        \caption{Fitch algorithm for binary trees. \\
          The unknown leaf node is discribed with both states. Computed internal nodes (exclusive the 
          root node) consists of three sets, where the last set is the final one (bold). \\
          From the second internal node (seen from the root node) there are several possibilities to 
          create the second and third set.}
        \label{fig: binary Fitch}
      \end{wrapfigure}
      First traverse from the leaf nodes to the root / move down the tree / postorder tree traversal.

        Each internal node is formed from its child nodes, where at the beginning the only information
        lies. \\
      Second traverse from the root node to the leafs. Each internal node is formed from its father node 
        and its sibling node. \\
      Last traversion (direction does not matter): Build the final state for every node. It is formed 
        from the sets of previous traversals. \\
      (The original Fitch algorithm was designed to minimize transitions without predicting actual states 
        of internal nodes, so it was just the first traversal.) \\
      The extension to the non-binary case is quite obvious, but holds some opportunities. In this case, 
        more than two children may be present for the first traversal, but the incision or union may 
        also be formed over more than two sets. Also in the second traversing, there may be several 
        sibling nodes. However, there are several possibilities here that were all tested and compared 
        in the simulation. Some of these options are already available in the binary case:
      \begin{itemize}
        \item The father node has (except for the root node) two state sets, because he came through 
          the up-traversing previously. Are both sets used or only the first traversing?
        \item Since there are several siblings, do \colorbox{red}{you} first of all make the cut or union, 
          or directly in the whole with the father node?
      \end{itemize}
      The first point already has an effect on the binary case. Figure \ref{fig: binary Fitch} shows 
        both possibilities of the three sets. \\
      Cunningham uses only the first state set of the father node \cite{Cunningham1998}. \\
      From these two points four different versions of Fitch were formed:
      \begin{enumerate}
        \item Fitch 1: First state set of father node; intersection/union of siblings first.
        \item Fitch 2: First state set of father node; intersection/union of siblings together with father node.
        \item Fitch 3: Both state sets of father node; intersection/union of siblings first.
        \item Fitch 4: Both state sets of father node; intersection/union of siblings together with father node sets.
      \end{enumerate}
      \begin{figure}
        \centering
        \includegraphics[width=0.8\textwidth]{Figures/simulation_fitch_evaluation.png}
        \caption{Test of Fitch Versions.}
        \label{fig:Fitch versions}
      \end{figure}
      These four versions were tested in the simulation with 100 trees and 10000 leaf nodes and a
        distribution of 60~\% FL to 40~\% P. Figure \ref{fig:Fitch versions} shows this over all unknown 
        node percentage. \\
      % At 90~\% unknown nodes and 90~\% of multifurcation of the internal nodes, version 1 was 89.26~\%, 
      %   version 2 was 89.26~\%, version 3 was 89.35~\%, and version 4 was 89.31~\% correct. Therefore, 
      %   only version 3 was used for all further simulations.
      At 95~\% unknown nodes and 95~\% of multifurcation of the internal nodes, version 1 was 88.37~\%, 
        version 2 was 88.37~\%, version 3 was 88.4~\%, and version 4 was 88.39~\% correct. Therefore, 
        only version 3 was used for all further simulations.

@@

    \begin{figure}[h!]
      \centering
      \includegraphics[width=1\textwidth]{Figures/Workflow-Simulation.png}
      \caption{A simulation was performed to compare different maximum parsimony algorithms. \\
        The course of the simulation with influence of the metadata analysis from the real data can
          be seen: \\
        (1) A phylogenetic tree is created randomly. (2) Simulate node states for all nodes. (3) 
          'Forget' internal states and some leaf node states. (4) 'Lose' phylogeny information. (5) 
          Make a nodelist for the algorithm. (6) Run algorithms. (7) Evaluate results. \\
        Points 3 and 4 are influenced by metadata of the real-data analysis.}
      \label{fig:Simulation Workflow}
    \end{figure}
    The course of a simulation is shown in Figure \ref{fig:Simulation Workflow}. The individual steps 
      are explained in the following. \\

    A tree is needed to do a simulation of ancestral state reconstruction. It had to be decided whether 
      to take the real tree or simulate a tree. In this simulation, trees are created randomly, as 
      one can replicate a complete binary phylogentic tree. Thus, there is also the possibility to 
      simulate the multifurcation. \\
    To get a random binary tree, the Phylo package from biopython is used \cite{Cock2009}. They offer 
      a randomized function which returns a BaseTree\footnote{
        \hyperlink{https://github.com/biopython/biopython/blob/master/Bio/Phylo/BaseTree.py}
        {https://github.com/biopython/biopython/blob/master/Bio/Phylo/BaseTree.py}
      }. \\
    \todo{ref in die discussion über die randomized function? Diskutieren wir das?}

    The next step is to simulate the states of the nodes using the transitions. Again, fully known 
      states are simulated and then everything is 'forgotten' but a few in the leaf nodes so that 
      \colorbox{red}{you} can later compare the reconstruction with the origin. \\
    The root node is defined as ancestor of all subsequent species and in this case, determined to be
      free-living. Therefore, a beta distribution for free-living is used at the beginning. Now 
      traverse from the root to the leaf nodes, always pulling out of the current distribution until 
      the randomly drawn number is above the threshold and the new node changes state. 
      \anmerkungstext{Für den Tree/Graph Traversal würde sich eine erklärende Abbildung anbieten. Nur 
      Text macht es sehr schwer verständlich an der Stelle. (Thilo)} \\
    To ensure that the parameter of the binomial distribution is restricted to the [0,1] interval, it 
      is modeled with a beta distribution as in Figure \ref{fig:Beta distribution}. \\

@@

% simulate loss of information
    Here begins the simulation of the lost information. This is on the one hand the states and on the 
      other the topology of the tree. \\

    In the real tree, there is usually only information about species living today $\rightarrow$ leaf 
      nodes. And beyond only a small percentage of these. All information about the states of the 
      internal node and one leaf node is 'forgotten' and stored in another column to the node. \\

    Last step for the preparation is the multifurcation of the tree. As previously explained, some 
      divisions in the tree are not known, so the real tree is not binary. This multifurcation is 
      simulated by an equally distributed percentage of forgotten internal nodes. \\

    Different percentages of forgetting information are simulated, see figure 
      \ref{fig:influence of unknown data} in the results. \\

    The last step is the evaluation of the results. This is done with a simple difference calculation 
      of the node states. \\
    In the nodelist, the originally simulated states and the newly calculated states are stored for 
      each node ($FL = 0$, $P = 1$). The sum of the differences of the node states gives the distance 
      of the prediction to the original tree.
